Supplement A: Community Clustering

### 2024-03-02 R/community-clustering.qmd
considerable refactoring to ensure everything is doing what it is supposed to be doing

## Introudction

@@

**Data:** VEPII and Cibola databases

@@

pretty_labels <- as_labeller(c(
  "n_communities" = "No. of Communities",
  "commute" = "Mean Commute to Nearest Center [mins]",
  "room_density" = "Room Density [N/sq.km]",
  "farm_area" = "Mean Area for Farms [sq.km/N]",
  "center_area" = "Mean Area for Centers [sq.km/N]",
  "farm_ratio" = "Ratio of Farms [N] to Centers [N]",
  "area" = "Total Area [sq.km]",
  "djoin" = "D-join [mins]",
  "dmax" = "D-max [mins]",
  "p" = "p [proportion in D-join]",
  "agg_factor" = "Grid Resolution [m]"
))

squished_labels <- as_labeller(c(
  "n_communities" = "No. of\nCommunities",
  "commute" = "Mean Commute\nto Nearest Center\n[mins]",
  "room_density" = "Room Density\n[N/sq.km]",
  "farm_area" = "Mean Area\nfor Farms [sq.km/N]",
  "center_area" = "Mean Area\nfor Centers [sq.km/N]",
  "farm_ratio" = "Ratio of\nFarms [N] to Centers [N]",
  "area" = "Total Area\n[sq.km]",
  "djoin" = "D-join\n[mins]",
  "dmax" = "D-max\n[mins]",
  "p" = "p\n[prop. in D-join]",
  "agg_factor" = "Grid Resolution\n[m]"
))

@@

Before converting the grid to a graph, we aggregate grid cells. Aggregating
reduces the resolution of the grid and by extension the number of grid cells the
raster contains. Since the grid cells are nodes in the grid graph, aggregating
simplifies the graph as well. This speeds up the path analysis considerably, but
it does come at the cost of precision in terms of path analysis across the
landscape, but that is not as bad as it sounds. The idea that we could build a
model that is perfectly isomorphic with the actual or true path taken by people
a thousand years ago is wildly implausible. The idea of the 'true path' is also
problematic as we're not evaluating a single commute, but rather the path people
should take more often than not. So, the real question is what level of
precision is sufficient for evaluating that. And the answer is, of course, it
just depends.

Are you walking blindfolded along the edge of a deep, precipitous cliff? Then
centimeters are probably important to you - really, really important to you. But
if you're just trying to get a reasonable estimate of the agricultural catchment
of an entirely pedestrian population based on the distribution of farms around
urban centers, probably the difference between \~30 m and \~80 m grid cells
isn't that big of a deal. If that doesn't convince you, though, we do some
limited sensitivity analysis down below.

@@

To calculate travel time between every pair of site points, we use a
bi-directional version of Dijkstra's algorithm as implemented in the R package
`{cppRouting}`. Note that we also average the from- and to- travel times, so the
results are isotropic (meaning direction is irrelevant).

@@

1.  Identify community centers
2.  Remove distant farms
3.  Join centers to their overlapping neighbors
4.  Join farms to their nearest center
5.  Draw smallest concave hull encompassing all farms, centers, and paths

Step 1 is already done and stored in the data. Step 2 can technically be done
after 2 and 3, but it makes sense to do it first, since it means those sites can
be ignored afterwards.

@@

@@

Here the catchment room count, $N$, of focal center $c$ is defined as

$$N_{c} = \sum_{k=1}^{S} R_{k} \cdot I(t_{kc} \leq D)$$

for all sites $S$ (including both farms and centers), with $R_{k}$ being the
room count at site $k$, $t_{kc}$ the travel time from $k$ to $c$, $D$ the
threshold travel time to a center (referred to as *D-join* in the paper), and
$I$ the indicator function that is 1 if $t \leq D$ and 0 otherwise. Note that
$S$ includes $c$ and that $t_{cc} = 0$ (the travel time from a center to the
same center is precisely 0), so $R_{c}$ is always included in $N_{c}$.

@@

$$N_{cd} = \sum_{k=1}^{S} R_{k} \cdot I(t_{kc} \leq D) \cdot I(t_{kd} \leq D)$$

@@

  # subset dm (distance matrix) to get dm[farms and centers, centers]
  center_cells <- x[["cell_table"]] |> filter(center == 1) |> pull(cell)

  j <- which(colnames(x[["dm"]]) %in% center_cells)

  dm <- x[["dm"]][, j]

  # find farms and centers that are within distance djoin of other centers
  djoins <- apply(dm, 1, \(.x){ which(.x <= djoin) })

  # remove farms that are isolated (ie, greater than djoin from a center)
  # note: this should just be farms because all centers are less than
  #   djoin from themselves
  djoins <- djoins[lengths(djoins) > 0]

  # collect list into a long-form table
  djoins <- tibble(
    center = colnames(dm)[unlist(djoins)],
    farm_or_center = rep(names(djoins), lengths(djoins))
  ) |> mutate(across(everything(), as.numeric))

  # now we combine the room counts with the djoins table
  djoins <- djoins |> 
    left_join(
      x[["cell_table"]] |> select(-center), 
      by = join_by(farm_or_center == cell)
    ) |> 
    rename("population" = n_room)

  # calculate the total population of the catchment around each center
  # note that this involves a lot of double-counting, farms and centers
  # counting towards the population size of multiple centers, but this is OK
  # as we are interested in using the shared population as a measure of 
  # interaction between centers, not as an estimate of a community's population
  gen_pop <- djoins |> 
    group_by(center) |> 
    summarize(population = sum(population))

  # for each center, what other centers have larger catchment populations?
  # using outer() is extremely fast
  # in this case, it creates an asymmetric logical matrix
  R <- outer(gen_pop[["population"]], gen_pop[["population"]], FUN = '<=')
  diag(R) <- FALSE

  colnames(R) <- gen_pop[["center"]]
  rownames(R) <- gen_pop[["center"]]

  # now, we turn that matrix into a big list 
  # each element is a vector with the cell IDs of the larger centers
  R <- apply(R, 1, which)
  R <- map(R, \(x){ as.numeric(names(x)) })

  # for each center, we look at all the centers with bigger catchment
  # populations than it, and ask: is the shared population size some fraction
  # of the total population of the catchment?
  neighbors_matrix <- purrr::imap(R, \(x, idx){

    N <- gen_pop |> 
      filter(center == as.numeric(idx)) |> 
      pull(population)

    focal_neighbors <- djoins |> 
      filter(center == as.numeric(idx)) |> 
      pull(farm_or_center)

    cntrs <- djoins |> 
      filter( # this filter gives the intersection of farms/cells 
        center %in% x,
        farm_or_center %in% c(idx, focal_neighbors)
      ) |> 
      group_by(center) |> 
      summarize(proportion = sum(population)/N) |> 
      filter(proportion >= p) |> 
      pull(center)

    # return a logical vector the length of R
    names(R) %in% cntrs

  neighbors_matrix <- do.call("rbind", neighbors_matrix)
  diag(neighbors_matrix) <- TRUE

  colnames(neighbors_matrix) <- names(R)
  rownames(neighbors_matrix) <- names(R)

  # there's a recursive dimension to this:
  # if a center is joined with a center that's joined with a center, etc.
  # i don't know how to do that in R
  # so, igraph it is...
  g <- igraph::graph_from_adjacency_matrix(neighbors_matrix)
  membership <- igraph::components(g)[["membership"]]

@@

@@

@@

Note that we also do some graph pruning in each iteration, basically cropping
the graph to an expanded area around the points assigned to a specific
community. This prevents the search algorithm from searching out away from the
other sites in the community.

@@

Here we show the community polygons themselves as well as the log-density of
rooms in each.

@@

ggsave(
  plot = gg,
  filename = fn,
  width = 6.7,
  height = 7.,
  dpi = 600
)

>>>>

## Introduction

@@

**Data:** Village Ecodynamics Project and Southwest Social Networks databases

@@

pretty_labels <- as_labeller(c(
  "n_communities" = "No. of Communities",
  "commute" = "Mean Commute to Nearest Center [mins]",
  "room_density" = "Room Density [N/sq.km]",
  "farm_area" = "Mean Area for Farms [sq.km/N]",
  "center_area" = "Mean Area for Centers [sq.km/N]",
  "farm_ratio" = "Ratio of Farms [N] to Centers [N]",
  "area" = "Total Area [sq.km]",
  "djoin" = "D-join [mins]",
  "dmax" = "D-max [mins]",
  "p" = "P [proportion in D-join]",
  "agg_factor" = "Grid Resolution [m]"
))

squished_labels <- as_labeller(c(
  "n_communities" = "No. of\nCommunities",
  "commute" = "Mean Commute\nto Nearest Center\n[mins]",
  "room_density" = "Room Density\n[N/sq.km]",
  "farm_area" = "Mean Area\nfor Farms [sq.km/N]",
  "center_area" = "Mean Area\nfor Centers [sq.km/N]",
  "farm_ratio" = "Ratio of\nFarms [N] to Centers [N]",
  "area" = "Total Area\n[sq.km]",
  "djoin" = "D-join\n[mins]",
  "dmax" = "D-max\n[mins]",
  "p" = "P\n[prop. in D-join]",
  "agg_factor" = "Grid Resolution\n[m]"
))

@@

Before converting the grid to a graph, we aggregate grid cells. Aggregating
reduces the resolution of the grid and by extension the number of grid cells the
raster contains. Since the grid cells are nodes in the grid graph, aggregating
simplifies the graph as well. This speeds up the path analysis considerably, but
it does come at the cost of precision in terms of path analysis across the
landscape, though that is not as bad as it sounds. The idea that we could build
a model that is perfectly isomorphic with the actual or true path taken by
people a thousand years ago is wildly implausible. The idea of the 'true path'
is also problematic as we're not evaluating a single commute, but rather the
path people should take more often than not. So, the real question is what level
of precision is sufficient for evaluating that. And the answer is, of course, it
just depends.

Are you walking blindfolded along the edge of a deep, precipitous cliff? Then
centimeters are probably important to you - really, really important to you. But
if you're just trying to get a reasonable estimate of the agricultural catchment
of an entirely pedestrian population based on the distribution of farms around
urban centers, probably the difference between \~30 m and \~90 m grid cells
isn't that big of a deal. If that doesn't convince you, though, we do some
limited sensitivity analysis down below.

@@

To calculate travel time between every pair of site points, we use Dijkstra's
algorithm as implemented in the R package `{cppRouting}`. Note that we also
average the from- and to- travel times, so the results are isotropic (meaning
direction is irrelevant).

@@

1.  Identify community centers
2.  Join centers to their overlapping neighbors
3.  Remove distant farms
4.  Join farms to their nearest center
5.  Draw smallest concave hull encompassing all farms, centers, and paths

Step 1 is already done and stored in the data. Step 3 is technically done before
2 since it means those sites can be ignored afterwards.

@@

    tibble(
      region = attr(x, "name"),
      n_outliers = attr(x, "n_outliers")
    )

  }) |> 
  bind_rows() |> 
  gt() |> 
  tab_header("No. of Dropped Farms") |> 
  opt_align_table_header("left") |> 
  tab_options(column_labels.hidden = TRUE)

@@

Here the catchment room count, $N$, of focal center $i$ is defined as

$$N_{c} = \sum_{i=1}^{S} R_{i} \cdot I(t_{ic} \leq D)$$

for all sites $S$ (including both farms and centers), with $R_{i}$ being the
room count at site $i$, $t_{ic}$ the travel time from $i$ to $c$, $D$ the
threshold travel time to a center (referred to as *D-join* in the paper), and
$I$ the indicator function that is 1 if $t \leq D$ and 0 otherwise. Note that
$S$ includes $c$ and that $t_{cc} = 0$ (the travel time from a center to the
same center is precisely 0), so $R_{c}$ is always included in $N_{c}$.

@@

$$N_{cd} = \sum_{i=1}^{S} R_{i} \cdot I(t_{ic} \leq D) \cdot I(t_{id} \leq D)$$

@@

  # subset dm (distance matrix) to get dm[farms and centers, centers]
  i <- x[["cell_table"]] |> pull(cell)
  j <- x[["cell_table"]] |> filter(center == 1) |> pull(cell)

  i <- which(rownames(x[["dm"]]) %in% i)
  j <- which(colnames(x[["dm"]]) %in% j)

  # room count in each farm/center
  Rk <- x[["cell_table"]] |> pull(n_room)
  names(Rk) <- x[["cell_table"]] |> pull(cell)

  # want to make sure these are arranged in the same order as rows in dm
  Rk <- Rk[rownames(dm)]

  # find farms and centers that are within distance djoin of other centers
  # this creates a binary/logical matrix of 1s and 0s,
  # which is what you get from the indicator function:
  # 1 if row is in distance djoin of column, 0 otherwise
  B <- matrix(as.numeric(dm <= djoin), nrow = nrow(dm))

  dimnames(B) <- dimnames(dm)

  # square community centers matrix
  G <- matrix(NA, nrow = ncol(dm), ncol = ncol(dm))

  colnames(G) <- colnames(dm)
  rownames(G) <- colnames(dm)

  # get total number of rooms within djoin of each center
  # this multiplies each column of B (which is just a bunch of 0s and 1s)
  # by the room count vector Rk, then takes the sum of the columns. 
  # we put this into the diagonal because it answers the question:
  # how many rooms does a center share with itself?
  # this is Nc. or Ncc since d = c.
  diag(G) <- colSums(B * Rk)

  # get shared room count (Ncd) within djoin of all pairs of centers
  # note we dump this into the lower triangle! this is a real R gotcha
  # as the vector is passed into the matrix by column, not row, so you 
  # can't do this with the upper triangle!
  lt <- lower.tri(G)

  G[lt] <- combn(
    ncol(B), 
    m = 2, 
    FUN = \(.x){ sum(B[, .x[1]] * B[, .x[2]] * Rk) }
  )

  # this matrix should be symmetric: 
  # the diagonal is Ncc, the room count a center shares with itself
  # the off-diagonals are Ncd = Ndc, the room counts shared between different centers
  ut <- upper.tri(G)

  G[ut] <- t(G)[ut]

  # create adjacency matrix
  # by dividing each column of G by the diagonal of G, we get the proportion
  # of the shared room count to the total room count, so p or pcd
  # we then compare this to the critical threshold, to determine adjacency
  neighbors_matrix <- (G / diag(G)) >= p

  # now to separate them into communities
  # unfortunately, there's a recursive dimension to this:
  # if a center is joined with a center that's joined with a center, etc.
  # i don't know how to do that in R, so igraph it is...
  g <- igraph::graph_from_adjacency_matrix(neighbors_matrix)
  membership <- igraph::components(g)[["membership"]]

@@

    tibble(
      region = attr(x, "name"),
      n_communities = attr(x, "n_communities")
    )

  }) |> 
  bind_rows() |> 
  gt() |> 
  tab_header("No. of Communities") |> 
  opt_align_table_header("left") |> 
  tab_options(column_labels.hidden = TRUE)

@@

    tibble(
      region = attr(x, "name"),
      n_dropped = attr(x, "n_dropped")
    )

  }) |> 
  bind_rows() |> 
  gt() |> 
  tab_header("No. of Dropped Communities") |> 
  opt_align_table_header("left") |> 
  tab_options(column_labels.hidden = TRUE)

@@

Note that we also do some graph pruning in each iteration, basically cropping
the graph to an expanded area around the points assigned to a specific
community. This prevents the search algorithm from searching out away from the
other sites in the community. The distance used for pruning is only meaningful
for pedestrian commute times.

@@

Here we show the community polygons themselves as well as the log-density of
rooms in each. Note the very pronounced edge effect along Mesa Verde National
Park's administrative boundary.

@@

ggsave(
  plot = gg,
  filename = fn,
  width = 7.1,
  height = 7.3,
  dpi = 600
)
